---
title: "EDA project: healthcare indicators in Utah"
---

---

## Overview

This project will be released on Thursday, June 6 and **conclude with a 10-minute presentation on Tuesday, June 18 during lecture time**.

Students will be randomly placed into groups of three and each group will be randomly assigned a dataset.

The goal of this project is to practice understanding the structure of a dataset, and to practice generating and evaluating hypotheses using fundamental EDA and data visualization techniques.

## Deliverables

Each group is expected to make slides to accompany the 10-minute presentation.

The presentation should feature the following:

-   Overview of the structure of your dataset

-   **Three questions/hypotheses** you are interested in exploring

-   **Three data visualizations** exploring the questions, at least two of which must be multivariate. **Each visualization must be in a different format** from the other two, and you must have at least one categorical and one continuous visualization

-   **One clustering analysis**

-   Conclusions reached for the hypotheses based on your EDA and data visualizations

## Timeline

There will be two submission deadlines:

**Thursday, June 13 at 5pm ET** - Each student will push their individual code for the project thus far to GitHub for review. We will then provide feedback.

**Monday, June 17 at 5pm ET** - Slides and full code must be completed and ready for presentation. Send your slides to Quang ([`quang@stat.cmu.edu`](mailto:quang@stat.cmu.edu)). All code must be written in `R`; but the slides may be created in any software. Take advantage of examples from lectures, but also feel free to explore online resources that may be relevant.

## Data

This dataset contains information on health indicators for 29 counties in Utah in 2014. The data are available online at the [State of Utah Open Data Catalog](https://opendata.utah.gov/) website. The provided dataset is a cleaned version of the original data (which can be found [here](https://opendata.utah.gov/Health/Health-Care-Indicators-By-Counties-In-Utah-2014/qmsu-gki4)).

Each row in the dataset corresponds to a county in Utah and the columns are:

-   `County`: name of the county in Utah
-   `Population`: county population
-   `PercentUnder18`: percent of county population that is under 18
-   `PercentOver65`: percent of county population that is 65 and over
-   `DiabeticRate`: diabetic rate
-   `HIVRate`: hive rate
-   `PrematureMortalityRate`: premature mortality rate
-   `InfantMortalityRate`: infant mortality rate
-   `ChildMortalityRate`: child mortality rate
-   `LimitedAccessToFood`: limited access to food %
-   `FoodInsecure`: % food insecure
-   `MotorDeathRate`: motor vehicle mortality rate
-   `DrugDeathRate`: drug poisoning mortality rate
-   `Uninsured`: % uninsured
-   `UninsuredChildren`: % uninsured children
-   `HealthCareCosts`: cost of healthcare
-   `CouldNotSeeDr`: % could not see doctor due to cost
-   `MedianIncome`: median household income
-   `ChildrenFreeLunch`: % children eligible for free lunch
-   `HomicideRate`: homicide rate

```{r, echo = FALSE, eval = FALSE}
# https://juliasilge.com/blog/health-care-indicators/
utah_health <- read_csv("https://opendata.utah.gov/resource/qmsu-gki4.csv")
utah_health <- utah_health |> 
  select(2:5, 18,22,24,27,31,34,36,38,42,44,48,51,55,60,63,64)
colnames(utah_health) <- c("County",
                           "Population",
                           "PercentUnder18",
                           "PercentOver65",
                           "DiabeticRate", 
                           "HIVRate",
                           "PrematureMortalityRate",
                           "InfantMortalityRate",
                           "ChildMortalityRate",
                           "LimitedAccessToFood",
                           "FoodInsecure", 
                           "MotorDeathRate",
                           "DrugDeathRate",
                           "Uninsured", 
                           "UninsuredChildren",
                           "HealthCareCosts", 
                           "CouldNotSeeDr",
                           "MedianIncome",
                           "ChildrenFreeLunch",
                           "HomicideRate")
write_csv(utah_health, "data/utah_health.csv")
```

## Starter code

```{r, eval=FALSE}
library(tidyverse)
utah_health <- read_csv("https://raw.githubusercontent.com/36-SURE/36-SURE.github.io/main/data/utah_health.csv")
```